---
title: Publication Surface Montage Reports from Statistical Maps
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Publication Surface Montage Reports from Statistical Maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

You have a set of second-level statistical volumes--for example signed
*t*-maps from several contrasts--and want one report that shows every map on an
inflated cortical surface with comparable thresholds, colour limits, anatomy,
and legends. A continuous map should read as a scalar field on cortex, not as
an atlas with parcel boundaries drawn over it. The *montage report* path keeps
those statistical and rendering semantics together. Its minimal recipe is two
calls:

```r
manifest <- build_manifest(source = ...)
render_montage_report(manifest, "report.html", surfatlas = atlas)
```

Everything else is policy and annotation. This vignette walks the whole path:
the manifest contract, `montage_policy()`, continuous volume-to-surface
projection, publication rendering, output formats, and the parcel/FDR wrinkles
you hit with ROI data.

```{r library}
library(neuromosaic)
```

## The render manifest

A render manifest is a data frame with **one row per statistical map**. It is
the single source of truth the report consumes; there is no hidden state.
`montage_manifest_schema()` returns the full column contract:

```{r schema}
schema <- montage_manifest_schema()
knitr::kable(schema[schema$required, c("field", "type", "role")],
             row.names = FALSE)
```

So the required columns are:

- **`map_id`** — a unique, stable key. It anchors caches, figure files, and
  table joins, so distinct maps must never share one.
- **`stat_kind`** — the statistic family, one of `t`, `z`, `beta`, `cope`.
- **`signed`** — a **logical** (note: not a string). Does the statistic carry
  positive *and* negative meaning? `TRUE` gives a symmetric diverging color
  scale; `FALSE` a one-sided scale.
- **`label`** — the human-facing panel title.

One more column is *conditionally* required: **`df`**. When a `t` map's
threshold is derived from a p-value (rather than supplied directly), the
degrees of freedom are needed for the p→*t* conversion. Supply `df`, or give an
explicit numeric `threshold`, for every `t` row.

The remaining schema columns (`path`, `recipe`, `p`, `threshold`, `tail`,
`level`, `n`, `description`, ...) are optional overrides; see
`?montage_manifest_schema`.

### Building one

When filenames do not follow BIDS conventions, the clearest route is an
explicit `source` data frame where you control every column:

```{r build}
manifest <- build_manifest(
  source = data.frame(
    map_id    = c("faces_gt_houses", "houses_gt_faces", "null_contrast"),
    path      = c("faces.nii.gz", "houses.nii.gz", "null.nii.gz"),
    stat_kind = "t",
    df        = 24,
    signed    = TRUE,
    label     = c("Faces > Houses", "Houses > Faces", "A null contrast"),
    p         = 0.001,
    stringsAsFactors = FALSE
  ),
  check_files = FALSE   # skip on-disk existence checks for this example
)
manifest[, c("map_id", "stat_kind", "signed", "df", "label")]
```

`build_manifest()` can also discover maps from a `pattern` glob or a path
vector and parse BIDS entities (`contrast-faces_model-m1_stat-z.nii.gz`); see
`?build_manifest`. However the maps are sourced, `validate_manifest()` enforces
the contract above before rendering.

## Thresholding and color policy

`montage_policy()` is the declarative knob set shared by every panel: default
p-value (and the p→threshold rule), tail, cluster connectivity, and the color
cap strategy. Per-map manifest columns override the policy where present.

```{r policy}
policy <- montage_policy(
  p    = 0.001,        # used when a row has no `threshold`/`p`
  tail = "two_sided"
)
policy$threshold      # NULL -> uses the built-in p->z/t rule
```

The built-in p→threshold rule covers `z` (via `qnorm`) and `t` (via `qt`, using
`df`). It is **uncorrected**.

### FDR / q-value thresholding

There is no `q =` option. FDR is data-dependent — the cutoff depends on a
map's whole p-value distribution — and the policy `threshold` function only
sees `(stat_kind, df, p, tail)`, not the voxel/parcel values, so it cannot
compute it. The supported pattern is to **precompute the FDR-critical
statistic value per map and put it in the manifest `threshold` column**, which
overrides the policy for that row. That is cleaner than zeroing
non-significant parcels in the volume and forcing a tiny `threshold` to bypass
the policy.

```{r fdr, eval = FALSE}
# Critical |t| at a Benjamini-Hochberg q from a map's p-values:
bh_threshold <- function(pvals, df, q = 0.05) {
  pvals <- sort(pvals[is.finite(pvals)])
  m <- length(pvals)
  ok <- pvals <= (seq_len(m) / m) * q
  # No survivors: a finite threshold above any statistic, so the map renders as
  # an empty panel (handled below). Avoid Inf, which the threshold check rejects.
  if (!any(ok)) return(.Machine$double.xmax)
  stats::qt(1 - pvals[max(which(ok))] / 2, df = df)
}

# map_pvalues: a list with one p-value vector per manifest row
manifest$threshold <- mapply(bh_threshold, map_pvalues, manifest$df, q = 0.05)
```

A map where nothing survives gets a threshold above any statistic, so it
becomes an empty panel — handled gracefully by default, as described next.

## Rendering surface montages

Surface panels switch on **when you pass a `surfatlas`** (internally
`render_surface = !is.null(surfatlas)`). `neuroatlas::get_schaefer_surfatlas()`
gives you a matching atlas:

```{r render, eval = FALSE}
atlas <- neuroatlas::get_schaefer_surfatlas(parcels = 400, networks = 17)

render_montage_report(
  manifest,
  output_file = "montage_report.html",
  surfatlas   = atlas,
  policy      = policy,
  image_dir   = "montage_images",   # where panel PNGs are written
  cache_dir   = "montage_cache"     # reused across re-renders
)
```

`image_dir` collects the generated PNGs; `cache_dir` lets repeated renders reuse
materialized maps and cached surface panels (keyed by map hash, threshold, and
cap, projection contract, atlas identity, mask/anatomy provenance, camera, and
other material rendering inputs) so iterating on labels or layout is cheap
without reusing a scientifically different panel. Supply a `bg` volume as well
to additionally generate volume montages, and an `atlas` to add per-panel peak
tables.

### What happens to a continuous volume?

For volumetric maps, `surf_montage()` selects the supported
`neuroatlas::plot_brain(style = "stat_publication")` path. The defaults are a
scientific contract rather than a collection of cosmetic switches:

- sample five fractions from 0.1 through 0.9 of the white-to-pial ribbon;
- use trilinear voxel interpolation and average across cortical depth;
- apply no tangential surface smoothing unless you request it in millimetres;
- interpolate the scalar within each visible triangle before thresholding and
  colour mapping;
- use a per-sample z-buffer, a cortex/medial-wall mask, and a subtle anatomy
  metric beneath the statistic;
- omit parcel, network, mesh, and internal occlusion lines; and
- derive the legend from the overlay limits, including signed endpoints, zero,
  and threshold marks.

The default renderer is a deterministic CPU backend, so the four-view path
does not need OpenGL, a browser, or the user's Chrome profile. The report
retains the effective projection, mask, anatomy, camera, colour, threshold, and
device settings in panel metadata.

You normally keep these defaults. When a methods section calls for an explicit
projection declaration, write it out through `surface_args`:

```{r projection-contract, eval=FALSE}
render_montage_report(
  manifest,
  "report.html",
  surfatlas = atlas,
  surface_args = list(
    sampling = "thickness",
    interpolation = "linear",
    aggregate = "mean",
    depth = seq(0.1, 0.9, length.out = 5),
    surface_smooth_fwhm = 0,
    medial_wall = "shade",
    camera = "canonical"
  )
)
```

Interpolation, aggregation across depth, and surface smoothing are independent
choices. In particular, the historical midpoint Gaussian-KNN path remains
available as `interpolation = "legacy"`, but it should not be described as
trilinear ribbon sampling. A categorical `aggregate = "mode"` is incompatible
with linear interpolation.

### What if the values really are parcel-level?

Parcel estimates remain a separate supported contract. Call `surf_montage()`
with `vals` instead of `stat`; those values retain atlas fill and boundary
semantics and are not silently routed through the continuous publication
preset.

```{r parcel-values, eval=FALSE}
surf_montage(
  vals = parcel_statistics,
  surfatlas = atlas,
  output_file = "parcel_map.png",
  threshold = 3,
  cap = 5
)
```

Keeping these paths separate prevents a continuous activation from acquiring
Schaefer borders merely because a Schaefer surface supplies its geometry and
cortex mask.

### Styling: `volume_args` / `surface_args` (and what's reserved)

Forward styling to the underlying `stat_montage()` / `surf_montage()` calls
through `volume_args` and `surface_args`:

```{r styling, eval = FALSE}
render_montage_report(
  manifest, "report.html", surfatlas = atlas,
  surface_args = list(
    overlay_alpha = 0.9,
    views = c("lateral", "medial"),
    camera = "presentation"
  )
)
```

The publication preset already uses overlay alpha `0.85`, canonical cameras,
and shaded medial wall by default. Use the slight presentation camera only as
an intentional display choice; keep canonical output when anatomical
orientation must be exact. Avoid recreating the preset with parcel-border,
silhouette, or mesh-line arguments--those lines have different scientific
meanings.

Some arguments are **renderer-managed and reserved** — passing them is an
error, because the report sets them per panel:

- `volume_args` reserves `bg`, `stat`, `draw`.
- `surface_args` reserves `stat`, `surfatlas`, `output_file`, `plot_fun`,
  `projection`.

Note these lists are merged with `modifyList()`, so passing `cap = NULL`
*removes* the derived default rather than forcing it; omit an argument to keep
the policy/robust default.

### Empty contrasts don't kill the report

In a group report it is common for one contrast (e.g. a null between-group
effect) to have no suprathreshold parcels. At the report level the default is
`empty = "warning"`: such a contrast renders its base surface with no overlay
and emits a warning, instead of aborting the whole run. Pass `empty = "error"`
to restore strict per-map behavior. (The lower-level `surf_montage()` /
`stat_montage()` keep their own `empty = "error"` default for direct callers.)

## Narrative context

A contrast name is rarely enough on its own. Four layers of author-written
markdown let you tell the reader what they are looking at, from the whole report
down to a single panel. All four accept markdown (bold, lists, links, `$math$`)
and are emitted verbatim.

- **Per panel** — the `description` manifest column (or a labeller's
  `description` field) prints under that panel's heading.
- **Report preamble** — `intro` prints once, before the maps.
- **Section narrative** — `section_notes` annotates a `layout` group.
- **Between maps** — `interludes` place free-standing text next to a panel,
  outside any panel's `description`.

```{r narrative, eval = FALSE}
render_montage_report(
  manifest,
  "report.html",
  surfatlas = atlas,
  layout    = c("model", "contrast"),

  # Report-level preamble, before any map.
  intro = "This report contrasts the **faces** and **houses** localizers across
           two GLM specifications. Warm colors are positive effects; see the QC
           table for dropped subjects.",

  # Section narrative keyed by layout columns. A row that fixes the first *k*
  # layout columns (deeper ones left NA) annotates that section's heading.
  section_notes = data.frame(
    model    = c("m1",              "m1"),
    contrast = c(NA,                "faces_gt_houses"),
    text     = c("Model 1: motion + WM/CSF nuisance regressors.",
                 "The canonical FFA contrast within model 1."),
    stringsAsFactors = FALSE
  ),

  # Free-standing text between panels, anchored to an existing map_id.
  interludes = data.frame(
    map_id   = "null_contrast",
    position = "before",           # or "after"
    text     = "The next map is a negative control: we expect no survivors.",
    stringsAsFactors = FALSE
  )
)
```

Inputs are validated against the resolved manifest, so an unknown `map_id`, a
`section_notes` column that is not a `layout` column, or a section path that
matches no map is an error rather than silently dropped text. On the CLI the
same three hooks are files: `--intro report-intro.md`, `--section-notes
sections.csv` (layout column(s) + a `text` column), and `--interludes
interludes.csv` (`map_id`, `text`, optional `position`).

## Output formats and portability

The extension of `output_file` selects the format:

- **`.html`** — the portable default. No external toolchain required.
- **`.pdf`** — renders through LaTeX, so it needs a TeX installation. The
  engine defaults to `latex_engine = "xelatex"` (common on workstations, often
  absent on minimal HPC nodes). Switch engines when needed:

  ```{r pdf, eval = FALSE}
  render_montage_report(manifest, "report.pdf", surfatlas = atlas,
                        latex_engine = "pdflatex")
  ```

- **`.qmd`** — does **not** render. It writes the Quarto source plus a
  `_report-data.rds` sidecar and prints a message telling you to run
  `quarto render` yourself. Use this when you want to hand-edit the report
  source before producing the final document.

If you are unsure whether your environment can produce a PDF, render `.html`
first; it always works.

Surface panels default to a 3000 by 1875 pixel, 300 dpi raster. PNG output
prefers `ragg` when available and otherwise uses the base PNG device. In PDF
reports, cortical panels remain rasterized while headings, orientation marks,
annotations, and legends are composed as vector-capable text and graphics.

## Recap

```r
manifest <- build_manifest(source = roi_table)        # the contract
policy   <- montage_policy(p = 0.001, tail = "two_sided")
atlas    <- neuroatlas::get_schaefer_surfatlas(400, 17)
render_montage_report(manifest, "report.html",        # volumes -> surfaces
                      surfatlas = atlas, policy = policy,
                      image_dir = "img", cache_dir = "cache")
```

See `?build_manifest`, `?montage_policy`, `?surf_montage`, and
`?render_montage_report` for the complete argument reference. For the
anatomical scene contract, continue with
`vignette("surface-panels", package = "neuroatlas")`; for the scalar-first
rasterizer and projection details, see
`vignette("surface-figures", package = "neurosurf")`.
