---
title: Grouped Montage Reports for Related Neuroimaging Maps
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Grouped Montage Reports for Related Neuroimaging Maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

You have one main result for each analysis--often a *z*- or *t*-statistic--and
several maps that help a reader interpret it: an effect estimate, its standard
error, perhaps a reliability diagnostic. You want one report where those maps
occupy the same slices or surface views, without forcing readers through four
nearly identical panels. A montage report groups associated maps and gives the
HTML reader a compact selector; PDF and print output expand every variant.

The minimal recipe is still two calls:

```r
manifest <- build_manifest(source = ...)
render_montage_report(manifest, "report.html", surfatlas = atlas)
```

Everything else is policy and annotation. This vignette shows how to name map
quantities, group variants, accept the display defaults, add a custom metric,
and render the same contract as slices or surfaces.

```{r library}
library(neuromosaic)
```

## The render manifest

A render manifest is a data frame with **one row per map variant**. It is
the single source of truth the report consumes; there is no hidden state.
`montage_manifest_schema()` returns the full column contract:

```{r schema}
schema <- montage_manifest_schema()
knitr::kable(schema[schema$required, c("field", "type", "role")],
             row.names = FALSE)
```

Every row needs:

- **`map_id`** — a unique, stable key. It anchors caches, figure files, and
  table joins, so distinct maps must never share one.
- **`quantity`** — what the values mean. Built-ins include
  `test_statistic`, `estimate`, `standard_error`, `variance`, `probability`,
  `proportion`, and `r_squared`; namespaced custom values are welcome.
- **`label`** — the human-facing panel title.

To associate variants, give them the same **`analysis_id`** and mark exactly one
as **`role = "primary"`**; the rest are `"auxiliary"`. A test statistic also
needs **`distribution = "z"`** or `"t"`. For a *t* statistic whose threshold is
derived from a p-value, supply `df`.

Old manifests remain valid: `stat_kind = "z"` or `"t"` maps to
`quantity = "test_statistic"`, while `"beta"` and `"cope"` map to
`"estimate"`. Those names are compatibility inputs, not the vocabulary custom
metrics must adopt. When `analysis_id` is absent, each row becomes its own
one-map analysis, so existing reports do not acquire selectors unexpectedly.

The remaining schema columns (`path`, `recipe`, `p`, `threshold`, `tail`,
`level`, `n`, `description`, ...) are optional overrides; see
`?montage_manifest_schema`.

### Group the maps from one analysis

When filenames do not follow BIDS conventions, the clearest route is an
explicit `source` data frame where you control every column:

```{r build}
manifest <- build_manifest(
  source = data.frame(
    analysis_id = "faces_gt_houses",
    map_id = c("faces_z", "faces_estimate", "faces_se", "faces_reliability"),
    path = c("faces_z.nii.gz", "faces_beta.nii.gz", "faces_se.nii.gz",
             "faces_reliability.nii.gz"),
    role = c("primary", "auxiliary", "auxiliary", "auxiliary"),
    quantity = c("test_statistic", "estimate", "standard_error",
                 "mylab:reliability"),
    distribution = c("z", NA, NA, NA),
    label = c("Z statistic", "Effect estimate", "Standard error",
              "Split-half reliability"),
    units = c("z", "% signal", "% signal", "correlation"),
    threshold = c(3.1, NA, NA, NA),
    stringsAsFactors = FALSE
  ),
  check_files = FALSE   # skip on-disk existence checks for this example
)
manifest[, c("analysis_id", "map_id", "role", "quantity", "label")]
```

```{r grouped-contract-check, include=FALSE}
stopifnot(
  length(unique(manifest$analysis_id)) == 1L,
  sum(manifest$role == "primary") == 1L,
  identical(manifest$map_id[[1L]], "faces_z")
)
```

`build_manifest()` can also discover maps from a `pattern` glob or a path
vector and parse BIDS entities (`contrast-faces_model-m1_stat-z.nii.gz`); see
`?build_manifest`. However the maps are sourced, `validate_manifest()` enforces
the contract above before rendering.

### Start from an ROI table

When the analysis result is already reduced to one row per atlas parcel, use
`parcel_render_manifest()` instead of writing one dense image per metric. The
adapter aligns every metric through `neuroatlas`, retains named parcel values
for surface rendering, and creates a `NeuroVol` in the atlas grid for slices,
peak tables, and optional interactive volume exploration:

```{r parcel-manifest, eval=FALSE}
atlas <- neuroatlas::get_schaefer_atlas(parcels = "100", networks = "7")
surface_atlas <- neuroatlas::get_schaefer_surfatlas(
  parcels = 100,
  networks = 7
)

roi_results <- data.frame(
  id = atlas$ids,
  z_stat = rnorm(length(atlas$ids)),
  beta = rnorm(length(atlas$ids)),
  standard_error = runif(length(atlas$ids), 0.05, 0.3),
  reliability = runif(length(atlas$ids), 0, 1)
)

manifest <- parcel_render_manifest(
  roi_results,
  atlas,
  metrics = c("z_stat", "beta", "standard_error", "reliability"),
  analysis_id = "faces_gt_houses"
)

render_montage_report(
  manifest,
  "parcel-report.html",
  bg = template_volume,
  atlas = atlas,
  surfatlas = surface_atlas,
  interactive = montage_interactive(),
  surface = montage_surface()
)
```

The first recognizable test-statistic column becomes primary. Common names
such as `z_stat`, `beta`, and `standard_error` receive the built-in display
profiles; an unknown numeric column becomes a namespaced custom quantity such
as `parcel:reliability`. Use `quantity`, `labels`, `units`, or `primary` to
override any inference. Strict full-atlas coverage is the default;
`allow_partial = TRUE` keeps missing parcels explicitly `NA`.

The ROI table remains the authoritative result. The parcel vector and volume
inside the manifest are display representations, so surface and volume panels
cannot silently disagree about row order or parcel identity.

## Let the quantity choose the display

You should not need to choose four palettes and four scales for a four-map
analysis. `resolve_montage_profiles()` applies conservative defaults:

- test statistics are thresholded with a diverging scale;
- estimates are continuous, zero-centred, and diverging;
- standard errors and variances are continuous, sequential, and nonnegative;
- probabilities, proportions, and R-squared maps use their fixed 0--1 domain;
- an unknown custom quantity is continuous and becomes diverging when its
  finite values include material negatives or its declared domain spans zero.

Robust limits use the 99th percentile rather than one extreme voxel. Auxiliary
maps use the full analysis mask by default; significance-masking them by the
primary map is available only as an explicit `support = "primary"` choice.

```{r profile-defaults}
map_values <- list(
  c(-5, -3, 0, 3, 5),
  c(-2.0, -0.4, 0, 0.5, 2.1),
  c(0.10, 0.14, 0.20, 0.31, 0.80),
  c(-0.4, 0.1, 0.5, 0.8)
)
resolved <- resolve_montage_profiles(manifest, map_values = map_values)
knitr::kable(
  resolved[, c("label", "effective_display_mode", "effective_scale",
               "effective_lower", "effective_upper")],
  digits = 2,
  row.names = FALSE
)
```

```{r profile-default-checks, include=FALSE}
stopifnot(
  identical(resolved$effective_display_mode,
            c("thresholded", "continuous", "continuous", "continuous")),
  identical(resolved$effective_scale,
            c("diverging", "diverging", "sequential", "diverging")),
  all(is.finite(resolved$effective_lower)),
  all(is.finite(resolved$effective_upper)),
  resolved$effective_lower[[3L]] >= 0
)
```

### Define a custom metric locally

A profile is an ordinary R object, not a global registration. Its legal domain
and units travel with the render call:

```{r custom-profile}
reliability_profile <- montage_map_profile(
  id = "mylab:reliability",
  label = "Split-half reliability",
  scale = "diverging",
  center = 0,
  domain = c(-1, 1),
  units = "correlation"
)

custom_resolved <- resolve_montage_profiles(
  manifest,
  profiles = list(`mylab:reliability` = reliability_profile),
  map_values = map_values
)
```

```{r custom-profile-checks, include=FALSE}
custom_row <- custom_resolved$quantity == "mylab:reliability"
stopifnot(
  identical(custom_resolved$effective_profile_source[custom_row], "user"),
  identical(custom_resolved$effective_lower[custom_row], -1),
  identical(custom_resolved$effective_upper[custom_row], 1)
)
```

The precedence is explicit manifest display columns, then a supplied profile,
then a built-in quantity profile, then automatic data-driven defaults. There is
no process-wide registry to make one report depend on another.

## Thresholding and color policy

`montage_policy()` is the declarative knob set shared by every panel: default
p-value (and the p→threshold rule), tail, cluster connectivity, and the color
cap strategy. Per-map manifest columns override the policy where present.

```{r policy}
policy <- montage_policy(
  p    = 0.001,        # used when a row has no `threshold`/`p`
  tail = "two_sided"
)
policy$threshold      # NULL -> uses the built-in p->z/t rule
```

The built-in p→threshold rule applies only to thresholded test-statistic rows.
It covers `z` (via `qnorm`) and `t` (via `qt`, using `df`) and is
**uncorrected**. Continuous auxiliary maps do not inherit that threshold.

### FDR / q-value thresholding

FDR is data-dependent: its cutoff depends on the whole map. The renderer can
resolve it per test-statistic map when you set `q`; continuous auxiliary maps
remain continuous.

```{r fdr, eval = FALSE}
policy <- montage_policy(q = 0.05)
render_montage_report(manifest, "report.html", policy = policy)
```

A map where nothing survives gets a threshold above any statistic, so it
becomes an empty panel — handled gracefully by default, as described next.

## Rendering surface montages

Surface panels switch on **when you pass a `surfatlas`** (internally
`render_surface = !is.null(surfatlas)`). `neuroatlas::get_schaefer_surfatlas()`
gives you a matching atlas:

```{r render, eval = FALSE}
atlas <- neuroatlas::get_schaefer_surfatlas(parcels = 400, networks = 17)

render_montage_report(
  manifest,
  output_file = "montage_report.html",
  surfatlas   = atlas,
  policy      = policy,
  profiles    = list(`mylab:reliability` = reliability_profile),
  map_selector = "auto",
  image_dir   = "montage_images",   # where panel PNGs are written
  cache_dir   = "montage_cache"     # reused across re-renders
)
```

`image_dir` collects the generated PNGs; `cache_dir` lets repeated renders reuse
materialized maps and cached surface panels (keyed by map hash, threshold, and
cap, projection contract, atlas identity, mask/anatomy provenance, camera, and
other material rendering inputs) so iterating on labels or layout is cheap
without reusing a scientifically different panel. Supply a `bg` volume as well
to additionally generate volume montages, and an `atlas` to add per-panel peak
tables.

### What does the HTML selector switch?

For two to four variants, `map_selector = "auto"` produces keyboard-accessible
segmented tabs. Larger groups use a select control. The primary map is selected
initially, and switching changes the whole variant: its slice and surface
images, description, metadata, QC warning, and table. A one-map analysis gets
no selector.

This is progressive enhancement. Before JavaScript runs--or when it is
disabled--all variants are visible. Printing and PDF output also show every
variant in primary-first order. Use `map_selector = "none"` when you prefer the
expanded HTML form.

### Add optional orthogonal exploration

The static panels are usually the right report artifact: they are quick to scan,
work in print, and remain usable on HPC systems where a browser cannot run on the
compute node. When a reader also needs to move through voxels, switch maps, or
try a different threshold or colour map, add `montage_interactive()` to the same
render call:

```{r interactive-volume, eval=FALSE}
render_montage_report(
  manifest,
  "report.html",
  bg = anatomical_volume,
  policy = policy,
  profiles = list(`mylab:reliability` = reliability_profile),
  interactive = montage_interactive(
    view = "orthogonal",
    assets = "embed",
    controls = c("threshold", "palette", "opacity")
  )
)
```

For a throw-away static website, change only the packaging choice and keep the
HTML beside its generated companion directory:

```{r interactive-bundle, eval=FALSE}
render_montage_report(
  manifest,
  "faces-site.html",
  bg = anatomical_volume,
  interactive = montage_interactive(assets = "bundle")
)
# Copy both faces-site.html and faces-site_files/, then serve that directory.
```

Nothing interactive loads until the reader chooses **Explore volume
interactively** under an analysis. The launcher opens three synchronized axial,
coronal, and sagittal views at the primary map's peak (or at the field-of-view
centre when requested). Its map menu follows the same primary-first order as the
static selector. Each map starts from the report-resolved display state:

- a thresholded test statistic gets its resolved tail and threshold;
- an estimate gets its continuous signed limits and diverging palette;
- a standard error or variance gets its nonnegative limits and sequential
  palette; and
- a namespaced custom metric follows its supplied or data-derived profile.

The threshold/range, colour-map, and opacity widgets are deliberately
exploratory. They update all three views, and **Reset to defaults** restores the
exact authored state for the currently selected map. They do not recompute
clusters, peaks, QC, tables, or the static figures, so those remain the
authoritative analysis result. A visible status marks a map as modified. Each
map keeps its own temporary settings while the page is open, and the static tabs
and interactive map menu remain synchronized. **Reset to report position** is
separate from the display reset and returns the crosshair to the authored world
coordinate.

Version 1 translates the resolved numeric limits, inclusive positive, negative,
or two-sided threshold, sequential/diverging palette family, units, and global
overlay alpha. neuroimjs exposes uniform layer opacity, so a static `soft` or
other intensity-dependent alpha mode is approximated by that global alpha. This
is a declared semantic-parity boundary: pixel identity is not claimed, and the
static montage is the authoritative rendering.

`assets = "embed"` places the gzip-compressed NIfTI payloads in the HTML and is
the convenient choice for a single file opened with `file://`. Embedded output
has an explicit `max_embed_mb` guard (25 MiB by default) applied to the actual
base64 payload, not just the smaller gzip files. `assets = "bundle"` writes a
complete `<report>_files/interactive/` tree containing `scene.json`, pinned
runtime and licence files, and content-addressed `.nii.gz` volumes. Bundle-mode
HTML deliberately uses companion resources rather than pretending to be a
single self-contained file; the HTML and directory must travel together and be
served over HTTP. It has no CDN or other network dependency. Compression is
lossless and does not crop, downsample, or quantize the maps.

Both modes include a prominent disclosure because the original voxel values
remain recoverable from the NIfTI payloads. Treat the resulting report as
sensitive data when its inputs are sensitive; gzip and base64 are packaging,
not anonymization or encryption.

The 25 MiB embed default is a guardrail, not a promise that every analysis fits.
`tools/benchmark-interactive-browser-fixtures.R` and
`tools/benchmark-interactive-browser.mjs` exercise the exact FLOAT64 NIfTI and
neuroimjs path for one background plus one, four, or eight overlays:

| Lattice and field | Raw, 8 maps | gzip, 1 / 4 / 8 maps | base64, 1 / 4 / 8 maps |
|:--|--:|--:|--:|
| 2 mm sparse | 61.98 MiB | 0.85 / 1.69 / 2.83 MiB | 1.13 / 2.26 / 3.77 MiB |
| 2 mm dense | 61.98 MiB | 5.43 / 12.55 / 22.04 MiB | 7.24 / 16.74 / 29.39 MiB |
| 1 mm sparse | 495.83 MiB | 6.76 / 13.52 / 22.58 MiB | 9.01 / 18.03 / 30.11 MiB |
| 1 mm dense | 495.83 MiB | 43.45 / 100.33 / 176.16 MiB | 57.93 / 133.78 / 234.88 MiB |

The same run measured fetch, gzip decode, NIfTI parse, and viewer update in the
project's headless Chromium. Times are wall-clock observations on the test host,
not universal performance promises:

| Lattice and field | Initial activation | Next 3 maps | Next 4 maps | Chromium backing store, 1 -> 4 -> 8 maps |
|:--|--:|--:|--:|--:|
| 2 mm sparse | 0.88 s | 0.37 s | 0.38 s | 14.22 -> 21.00 -> 21.00 MiB |
| 2 mm dense | 0.80 s | 0.33 s | 0.42 s | 14.22 -> 21.00 -> 21.00 MiB |
| 1 mm sparse | 1.17 s | 0.87 s | 0.93 s | 112.78 -> 167.01 -> 167.01 MiB |
| 1 mm dense | 1.46 s | 1.74 s | 1.90 s | 112.78 -> 167.01 -> 167.01 MiB |

These deterministic measurements were recorded on 2026-08-22. Chromium's JS
heap increase stayed between 4.38 and 5.13 MiB; the backing-store measure covers
decoded buffers but not driver-specific GPU memory. Actual values and masks can
compress differently. The evidence supports embedding a typical small 2 mm
group while making dense 1 mm groups choose `assets = "bundle"` or an explicitly
reviewed higher budget. The plateau from four to eight maps verifies the default
two-overlay cache bound; the background and selected map still decode first.

The browser path is fail-closed: background and overlays must agree in dimensions,
voxel spacing, orientation, origin, and full affine. It never restamps, resamples,
or silently aligns data, and `volume_args = list(on_mismatch = "restamp")` is
therefore rejected when interactivity is requested. If JavaScript, WebGL, data
loading, or geometry validation fails, the report leaves the static montage in
place with a local error message. PDF output always remains static; QMD output
retains the scene and companion assets for a later HTML render. Its render
sidecar omits source paths, recipes, and redundant map objects, and rewrites or
copies static panel images under the companion directory so the QMD, sidecar,
and `<report>_files/` tree can move together.

For troubleshooting, a bundle opened with `file://` should be served from a
local or static HTTP server, or regenerated with `assets = "embed"`. A geometry
error must be repaired upstream by aligning inputs; the interactive path never
restamps. A WebGL error affects only the optional canvas. If controls show the
modified status, **Reset to defaults** and **Reset to report position** restore
the authored display and coordinate independently. Hiding or leaving the page
disposes the viewer and its bounded decoded-map cache; reopening starts from
the currently selected map's report defaults. Report provenance records
the VolumeScene version, neuroimjs and adapter versions, runtime SHA-256, asset
hashes, compression, byte totals, and declared storage transformations so a
copied artifact can be audited without its source checkout.

Custom HTML templates can use the same stable emitters without constructing an
htmlwidget:

```{r interactive-hooks, eval=FALSE}
hooks <- montage_interactive_report_hooks(
  report_data$interactive,
  is_html = knitr::is_html_output()
)
hooks$emit_document()                 # exactly once
hooks$emit_host("faces-versus-places") # after that analysis's static panels
```

An htmlwidget wrapper is unnecessary for the report path: the scene, runtime,
and payload lifecycle belong to the report document. Applications that need
Shiny-style reactive bindings can add a separate wrapper around the same scene
contract later.

### What happens to a continuous volume?

For volumetric maps, `surf_montage()` selects the supported
`neuroatlas::plot_brain(style = "stat_publication")` path. The defaults are a
scientific contract rather than a collection of cosmetic switches:

- sample five fractions from 0.1 through 0.9 of the white-to-pial ribbon;
- use trilinear voxel interpolation and average across cortical depth;
- apply no tangential surface smoothing unless you request it in millimetres;
- interpolate the scalar within each visible triangle before thresholding and
  colour mapping;
- use a per-sample z-buffer, a cortex/medial-wall mask, and a subtle anatomy
  metric beneath the statistic;
- omit parcel, network, mesh, and internal occlusion lines; and
- derive the legend from the overlay limits, including signed endpoints and
  zero, plus threshold marks when the map is thresholded.

Within an analysis, the primary map chooses the slice coordinates; surface
variants share the same views and camera. A requested shared color scale is
used only when quantity, resolved profile, units, scale type, and center are
compatible.

The default renderer is a deterministic CPU backend, so the four-view path
does not need OpenGL, a browser, or the user's Chrome profile. The report
retains the effective projection, mask, anatomy, camera, colour, threshold, and
device settings in panel metadata.

You normally keep these defaults. When a methods section calls for an explicit
projection declaration, write it out through `surface_args`:

```{r projection-contract, eval=FALSE}
render_montage_report(
  manifest,
  "report.html",
  surfatlas = atlas,
  surface_args = list(
    sampling = "thickness",
    interpolation = "linear",
    aggregate = "mean",
    depth = seq(0.1, 0.9, length.out = 5),
    surface_smooth_fwhm = 0,
    medial_wall = "shade",
    camera = "canonical"
  )
)
```

Interpolation, aggregation across depth, and surface smoothing are independent
choices. In particular, the historical midpoint Gaussian-KNN path remains
available as `interpolation = "legacy"`, but it should not be described as
trilinear ribbon sampling. A categorical `aggregate = "mode"` is incompatible
with linear interpolation.

### What if the values really are parcel-level?

Parcel estimates remain a separate supported contract. Call `surf_montage()`
with `vals` instead of `stat`; those values retain atlas fill and boundary
semantics and are not silently routed through the continuous publication
preset.

```{r parcel-values, eval=FALSE}
surf_montage(
  vals = parcel_statistics,
  surfatlas = atlas,
  output_file = "parcel_map.png",
  threshold = 3,
  cap = 5
)
```

Keeping these paths separate prevents a continuous activation from acquiring
Schaefer borders merely because a Schaefer surface supplies its geometry and
cortex mask.

### Styling: `volume_args` / `surface_args` (and what's reserved)

Forward styling to the underlying `stat_montage()` / `surf_montage()` calls
through `volume_args` and `surface_args`:

```{r styling, eval = FALSE}
render_montage_report(
  manifest, "report.html", surfatlas = atlas,
  surface_args = list(
    overlay_alpha = 0.9,
    views = c("lateral", "medial"),
    camera = "presentation"
  )
)
```

The publication preset already uses overlay alpha `0.85`, canonical cameras,
and shaded medial wall by default. Use the slight presentation camera only as
an intentional display choice; keep canonical output when anatomical
orientation must be exact. Avoid recreating the preset with parcel-border,
silhouette, or mesh-line arguments--those lines have different scientific
meanings.

Some arguments are **renderer-managed and reserved** — passing them is an
error, because the report sets them per panel:

- `volume_args` reserves `bg`, `stat`, `draw`.
- `surface_args` reserves `stat`, `surfatlas`, `output_file`, `plot_fun`,
  `projection`.

Note these lists are merged with `modifyList()`, so passing `cap = NULL`
*removes* the derived default rather than forcing it; omit an argument to keep
the policy/robust default.

### Empty contrasts don't kill the report

In a group report it is common for one contrast (e.g. a null between-group
effect) to have no suprathreshold parcels. At the report level the default is
`empty = "warning"`: such a contrast renders its base surface with no overlay
and emits a warning, instead of aborting the whole run. Pass `empty = "error"`
to restore strict per-map behavior. (The lower-level `surf_montage()` /
`stat_montage()` keep their own `empty = "error"` default for direct callers.)

## Narrative context

A contrast name is rarely enough on its own. Four layers of author-written
markdown let you tell the reader what they are looking at, from the whole report
down to a single panel. All four accept markdown (bold, lists, links, `$math$`)
and are emitted verbatim.

- **Per panel** — the `description` manifest column (or a labeller's
  `description` field) prints under that panel's heading.
- **Report preamble** — `intro` prints once, before the maps.
- **Section narrative** — `section_notes` annotates a `layout` group.
- **Between maps** — `interludes` place free-standing text next to a panel,
  outside any panel's `description`.

```{r narrative, eval = FALSE}
render_montage_report(
  manifest,
  "report.html",
  surfatlas = atlas,
  layout    = c("model", "contrast"),

  # Report-level preamble, before any map.
  intro = "This report contrasts the **faces** and **houses** localizers across
           two GLM specifications. Warm colors are positive effects; see the QC
           table for dropped subjects.",

  # Section narrative keyed by layout columns. A row that fixes the first *k*
  # layout columns (deeper ones left NA) annotates that section's heading.
  section_notes = data.frame(
    model    = c("m1",              "m1"),
    contrast = c(NA,                "faces_gt_houses"),
    text     = c("Model 1: motion + WM/CSF nuisance regressors.",
                 "The canonical FFA contrast within model 1."),
    stringsAsFactors = FALSE
  ),

  # Free-standing text between panels, anchored to an existing map_id.
  interludes = data.frame(
    map_id   = "null_contrast",
    position = "before",           # or "after"
    text     = "The next map is a negative control: we expect no survivors.",
    stringsAsFactors = FALSE
  )
)
```

Inputs are validated against the resolved manifest, so an unknown `map_id`, a
`section_notes` column that is not a `layout` column, or a section path that
matches no map is an error rather than silently dropped text. On the CLI the
same three hooks are files: `--intro report-intro.md`, `--section-notes
sections.csv` (layout column(s) + a `text` column), and `--interludes
interludes.csv` (`map_id`, `text`, optional `position`).

## Output formats and portability

The extension of `output_file` selects the format:

- **`.html`** — the portable default. No external toolchain required.
- **`.pdf`** — renders through LaTeX, so it needs a TeX installation. The
  engine defaults to `latex_engine = "xelatex"` (common on workstations, often
  absent on minimal HPC nodes). Switch engines when needed:

  ```{r pdf, eval = FALSE}
  render_montage_report(manifest, "report.pdf", surfatlas = atlas,
                        latex_engine = "pdflatex")
  ```

- **`.qmd`** — does **not** render. It writes the Quarto source plus a
  `_report-data.rds` sidecar and prints a message telling you to run
  `quarto render` yourself. Use this when you want to hand-edit the report
  source before producing the final document.

If you are unsure whether your environment can produce a PDF, render `.html`
first; it always works.

Surface panels default to a 3000 by 1875 pixel, 300 dpi raster. PNG output
prefers `ragg` when available and otherwise uses the base PNG device. In PDF
reports, cortical panels remain rasterized while headings, orientation marks,
annotations, and legends are composed as vector-capable text and graphics.

## Recap

```r
manifest <- build_manifest(source = grouped_maps)     # one row per variant
policy   <- montage_policy(p = 0.001, tail = "two_sided")
atlas    <- neuroatlas::get_schaefer_surfatlas(400, 17)
render_montage_report(manifest, "report.html",        # volumes -> surfaces
                      surfatlas = atlas, policy = policy,
                      profiles = list(`mylab:reliability` = reliability_profile),
                      image_dir = "img", cache_dir = "cache")
```

See `?build_manifest`, `?montage_map_profile`, `?montage_policy`,
`?surf_montage`, and
`?render_montage_report` for the complete argument reference. For the
anatomical scene contract, continue with
`vignette("surface-panels", package = "neuroatlas")`; for the scalar-first
rasterizer and projection details, see
`vignette("surface-figures", package = "neurosurf")`.
