---
title: "Surface Montage Reports from ROI/Parcel Statistical Maps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Surface Montage Reports from ROI/Parcel Statistical Maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

You have a set of second-level statistical maps — say per-parcel *t*-statistics
from a Schaefer-400/17 ROI cross-decoding analysis — and you want a single
report that shows each contrast on an inflated cortical surface, with consistent
thresholding and color scales. The *montage report* path does this. Its minimal
recipe is two calls:

```r
manifest <- build_manifest(source = ...)
render_montage_report(manifest, "report.html", surfatlas = atlas)
```

Everything else is policy and styling. This vignette walks the whole path:
the manifest contract, `montage_policy()`, surface rendering, output formats,
and the parcel/FDR wrinkles you hit with ROI data.

```{r library}
library(neuromosaic)
```

## The render manifest

A render manifest is a data frame with **one row per statistical map**. It is
the single source of truth the report consumes; there is no hidden state.
`montage_manifest_schema()` returns the full column contract:

```{r schema}
schema <- montage_manifest_schema()
knitr::kable(schema[schema$required, c("field", "type", "role")],
             row.names = FALSE)
```

So the required columns are:

- **`map_id`** — a unique, stable key. It anchors caches, figure files, and
  table joins, so distinct maps must never share one.
- **`stat_kind`** — the statistic family, one of `t`, `z`, `beta`, `cope`.
- **`signed`** — a **logical** (note: not a string). Does the statistic carry
  positive *and* negative meaning? `TRUE` gives a symmetric diverging color
  scale; `FALSE` a one-sided scale.
- **`label`** — the human-facing panel title.

One more column is *conditionally* required: **`df`**. When a `t` map's
threshold is derived from a p-value (rather than supplied directly), the
degrees of freedom are needed for the p→*t* conversion. Supply `df`, or give an
explicit numeric `threshold`, for every `t` row.

The remaining schema columns (`path`, `recipe`, `p`, `threshold`, `tail`,
`level`, `n`, `description`, ...) are optional overrides; see
`?montage_manifest_schema`.

### Building one

For ROI/atlas work the filenames rarely follow BIDS conventions, so the
clearest route is an explicit `source` data frame where you control every
column:

```{r build}
manifest <- build_manifest(
  source = data.frame(
    map_id    = c("faces_gt_houses", "houses_gt_faces", "null_contrast"),
    path      = c("faces.nii.gz", "houses.nii.gz", "null.nii.gz"),
    stat_kind = "t",
    df        = 24,
    signed    = TRUE,
    label     = c("Faces > Houses", "Houses > Faces", "A null contrast"),
    p         = 0.001,
    stringsAsFactors = FALSE
  ),
  check_files = FALSE   # skip on-disk existence checks for this example
)
manifest[, c("map_id", "stat_kind", "signed", "df", "label")]
```

`build_manifest()` can also discover maps from a `pattern` glob or a path
vector and parse BIDS entities (`contrast-faces_model-m1_stat-z.nii.gz`); see
`?build_manifest`. However the maps are sourced, `validate_manifest()` enforces
the contract above before rendering.

## Thresholding and color policy

`montage_policy()` is the declarative knob set shared by every panel: default
p-value (and the p→threshold rule), tail, cluster connectivity, and the color
cap strategy. Per-map manifest columns override the policy where present.

```{r policy}
policy <- montage_policy(
  p    = 0.001,        # used when a row has no `threshold`/`p`
  tail = "two_sided"
)
policy$threshold      # NULL -> uses the built-in p->z/t rule
```

The built-in p→threshold rule covers `z` (via `qnorm`) and `t` (via `qt`, using
`df`). It is **uncorrected**.

### FDR / q-value thresholding

There is no `q =` option. FDR is data-dependent — the cutoff depends on a
map's whole p-value distribution — and the policy `threshold` function only
sees `(stat_kind, df, p, tail)`, not the voxel/parcel values, so it cannot
compute it. The supported pattern is to **precompute the FDR-critical
statistic value per map and put it in the manifest `threshold` column**, which
overrides the policy for that row. That is cleaner than zeroing
non-significant parcels in the volume and forcing a tiny `threshold` to bypass
the policy.

```{r fdr, eval = FALSE}
# Critical |t| at a Benjamini-Hochberg q from a map's p-values:
bh_threshold <- function(pvals, df, q = 0.05) {
  pvals <- sort(pvals[is.finite(pvals)])
  m <- length(pvals)
  ok <- pvals <= (seq_len(m) / m) * q
  if (!any(ok)) return(Inf)                 # nothing survives -> empty map
  stats::qt(1 - pvals[max(which(ok))] / 2, df = df)
}

# map_pvalues: a list with one p-value vector per manifest row
manifest$threshold <- mapply(bh_threshold, map_pvalues, manifest$df, q = 0.05)
```

A row whose `threshold` is `Inf` (nothing survives) becomes an empty panel —
handled gracefully by default, as described next.

## Rendering surface montages

Surface panels switch on **when you pass a `surfatlas`** (internally
`render_surface = !is.null(surfatlas)`). `neuroatlas::get_schaefer_surfatlas()`
gives you a matching atlas:

```{r render, eval = FALSE}
atlas <- neuroatlas::get_schaefer_surfatlas(parcels = 400, networks = 17)

render_montage_report(
  manifest,
  output_file = "montage_report.html",
  surfatlas   = atlas,
  policy      = policy,
  image_dir   = "montage_images",   # where panel PNGs are written
  cache_dir   = "montage_cache"     # reused across re-renders
)
```

`image_dir` collects the generated PNGs; `cache_dir` lets repeated renders reuse
materialized maps and cached surface panels (keyed by map hash, threshold, and
cap) so iterating on labels or layout is cheap. Supply a `bg` volume as well to
additionally generate volume montages, and an `atlas` to add per-panel peak
tables.

### Styling: `volume_args` / `surface_args` (and what's reserved)

Forward styling to the underlying `stat_montage()` / `surf_montage()` calls
through `volume_args` and `surface_args`:

```{r styling, eval = FALSE}
render_montage_report(
  manifest, "report.html", surfatlas = atlas,
  surface_args = list(overlay_alpha = 0.9, views = c("lateral", "medial"))
)
```

Some arguments are **renderer-managed and reserved** — passing them is an
error, because the report sets them per panel:

- `volume_args` reserves `bg`, `stat`, `draw`.
- `surface_args` reserves `stat`, `surfatlas`, `output_file`, `plot_fun`,
  `projection`.

Note these lists are merged with `modifyList()`, so passing `cap = NULL`
*removes* the derived default rather than forcing it; omit an argument to keep
the policy/robust default.

### Empty contrasts don't kill the report

In a group report it is common for one contrast (e.g. a null between-group
effect) to have no suprathreshold parcels. At the report level the default is
`empty = "warning"`: such a contrast renders its base surface with no overlay
and emits a warning, instead of aborting the whole run. Pass `empty = "error"`
to restore strict per-map behavior. (The lower-level `surf_montage()` /
`stat_montage()` keep their own `empty = "error"` default for direct callers.)

## Output formats and portability

The extension of `output_file` selects the format:

- **`.html`** — the portable default. No external toolchain required.
- **`.pdf`** — renders through LaTeX, so it needs a TeX installation. The
  engine defaults to `latex_engine = "xelatex"` (common on workstations, often
  absent on minimal HPC nodes). Switch engines when needed:

  ```{r pdf, eval = FALSE}
  render_montage_report(manifest, "report.pdf", surfatlas = atlas,
                        latex_engine = "pdflatex")
  ```

- **`.qmd`** — does **not** render. It writes the Quarto source plus a
  `_report-data.rds` sidecar and prints a message telling you to run
  `quarto render` yourself. Use this when you want to hand-edit the report
  source before producing the final document.

If you are unsure whether your environment can produce a PDF, render `.html`
first; it always works.

## Recap

```r
manifest <- build_manifest(source = roi_table)        # the contract
policy   <- montage_policy(p = 0.001, tail = "two_sided")
atlas    <- neuroatlas::get_schaefer_surfatlas(400, 17)
render_montage_report(manifest, "report.html",        # surfatlas -> surfaces
                      surfatlas = atlas, policy = policy,
                      image_dir = "img", cache_dir = "cache")
```

See `?build_manifest`, `?montage_policy`, and `?render_montage_report` for the
complete argument reference.
