---
title: "Exploring Ad Hoc Neurotabs Datasets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring Ad Hoc Neurotabs Datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)

has_neurotabs <- requireNamespace("neurotabs", quietly = TRUE)
has_explorer_deps <- has_neurotabs &&
  requireNamespace("shiny", quietly = TRUE) &&
  requireNamespace("DT", quietly = TRUE) &&
  requireNamespace("ggiraph", quietly = TRUE)
```

Suppose you have one subject-level map per folder:

- `sub-01/maps/AUC.nii.gz`
- `sub-02/maps/AUC.nii.gz`
- `...`
- `sub-30/maps/AUC.nii.gz`

and a TSV file with `subject`, a continuous covariate `measure`, and a
two-level grouping variable `group`. You do not want to write a neurotabs
manifest by hand. You want to build an `nftab` directly from that table and
open the interactive explorer so that, after picking a cluster or parcel, the
signal plot shows `AUC ~ measure` with separate fitted lines for each `group`.

That workflow is:

1. Build an ad hoc `nftab` with `neurotabs::nf_from_table()`.
2. Pass that dataset to `nf_cluster_explorer()`.
3. In the app, set `Plot Variable = measure` and `Color / Group = group`.

`clusterreport::nf_cluster_explorer()` uses `neurotabs::nf_design(ds)`
automatically, so the columns in your TSV become explorer controls without a
separate `design` argument.

## Create a nested AUC dataset

This example creates a temporary study directory with the same nested structure
you would have on disk. The group-level `stat_map` and atlas come from
`example_cluster_inputs()`, while the subject-level AUC volumes are written as
separate NIfTI files.

```{r make-observation-table}
inputs <- clusterreport::example_cluster_inputs()
n_subjects <- 30L

obs <- tibble::tibble(
  subject = sprintf("sub-%02d", seq_len(n_subjects)),
  group = rep(c("control", "patient"), each = n_subjects / 2),
  measure = seq(-1.5, 1.5, length.out = n_subjects),
  auc_path = file.path(subject, "maps", "AUC.nii.gz")
)

head(obs, 4)
```

```{r write-auc-volumes, eval = has_neurotabs, include = FALSE}
study_dir <- file.path(tempdir(), "clusterreport-auc-study")
dir.create(study_dir, recursive = TRUE, showWarnings = FALSE)

space_3d <- neuroim2::space(inputs$stat_map)
dims <- as.integer(dim(inputs$stat_map))[1:3]

for (i in seq_len(n_subjects)) {
  subject_dir <- file.path(study_dir, obs$subject[i], "maps")
  dir.create(subject_dir, recursive = TRUE, showWarnings = FALSE)

  arr <- array(stats::rnorm(prod(dims), sd = 0.05), dim = dims)
  arr[2:4, 2:4, 2:4] <- 0.4 + 0.9 * obs$measure[i] +
    ifelse(obs$group[i] == "patient", 0.6, -0.6) +
    stats::rnorm(27, sd = 0.08)
  arr[7:9, 7:9, 7:9] <- 0.2 - 0.5 * obs$measure[i] +
    stats::rnorm(27, sd = 0.08)

  neuroim2::write_vol(
    neuroim2::NeuroVol(arr, space = space_3d),
    file.path(study_dir, obs$auc_path[i])
  )
}

obs_path <- file.path(study_dir, "participants.tsv")
utils::write.table(obs, obs_path, sep = "\t", quote = FALSE, row.names = FALSE)

stopifnot(file.exists(obs_path))
stopifnot(all(file.exists(file.path(study_dir, obs$auc_path))))
```

## Build the `nftab` without a manifest

`nf_from_table()` is the ad hoc entry point in neurotabs. The `auc_path` column
stores relative file paths, and `root` tells neurotabs where those files live.
Here `axes = "subject"` keeps `group` and `measure` as row metadata rather than
observation axes.

```{r build-nftab, eval = has_neurotabs}
ds <- neurotabs::nf_from_table(
  observations = obs_path,
  feature = "AUC",
  locator_col = "auc_path",
  axes = "subject",
  space = "MNI152NLin2009cAsym",
  dataset_id = "auc-demo",
  root = study_dir
)

ds
```

```{r inspect-design, eval = has_neurotabs}
design_tbl <- neurotabs::nf_design(ds)
design_tbl[1:4, c("subject", "group", "measure")]
```

Those design columns are the bridge into `clusterreport`:
`clusterreport::nf_cluster_explorer()` reads `nf_design(ds)` and merges it into
the extracted cluster signal table.

## Launch the interactive explorer

`cluster_explorer()` expects both a volumetric atlas and a `surfatlas`. If you
already have a real surface atlas, use it. For a volume-only workflow, a small
label-only `surfatlas` is enough to enable the fallback parcel view used here.

```{r make-surfatlas}
surfatlas <- structure(
  list(
    ids = inputs$atlas$ids,
    labels = inputs$atlas$labels,
    hemi = inputs$atlas$hemi
  ),
  class = c("toy_surf", "surfatlas", "atlas")
)
```

```{r build-app, eval = has_explorer_deps}
app <- clusterreport::nf_cluster_explorer(
  ds = ds,
  stat_map = inputs$stat_map,
  data_feature = "AUC",
  atlas = inputs$atlas,
  surfatlas = surfatlas,
  threshold = 3,
  min_cluster_size = 5
)

inherits(app, "shiny.appobj")
```

In an interactive R session, run `app` or `shiny::runApp(app)`.

Once the app is open:

1. Click a parcel in the brain view or select a cluster from the table.
2. Set `Plot Variable` to `measure`.
3. Set `Color / Group` to `group`.
4. Leave `Collapse By` empty unless you want to average repeated rows first.

The signal panel will show subject-level AUC values for the selected cluster,
with points colored by `group` and separate linear fits overlaid for each
group. If you select multiple clusters, the explorer facets them so each panel
keeps the same `measure`/`group` mapping.

## Direct template for your own study

If your data already lives on disk, the only part you need to change is the
table and paths:

```{r your-study-template, eval = FALSE}
obs <- utils::read.delim("~/study/participants.tsv")

ds <- neurotabs::nf_from_table(
  observations = obs,
  feature = "AUC",
  locator_col = "auc_path",
  axes = "subject",
  space = "MNI152NLin2009cAsym",
  root = "~/study"
)

app <- clusterreport::nf_cluster_explorer(
  ds = ds,
  stat_map = stat_map,
  data_feature = "AUC",
  atlas = atlas,
  surfatlas = surfatlas
)
```

For larger datasets, `nf_cluster_explorer(..., strategy = "lazy")` avoids
stacking every subject volume up front and samples from the NIfTI files on
demand.
