---
title: Generating Cluster Reports from Neuroimaging Results
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Generating Cluster Reports from Neuroimaging Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

You have a statistical map from a group analysis, a 4-D volume of betas for
each condition and time point, and a design table describing what each image
represents. You want a PDF report with brain images, MNI coordinate tables, and
condition-by-time plots for every significant cluster. `cluster_report()` does
all of that in a single call.

## Quick start

```{r quick-start}
library(neuromosaic)

inputs <- example_cluster_inputs()
result <- cluster_report(
  stat_map    = inputs$stat_map,
  data_source = inputs$data_source,
  atlas       = inputs$atlas,
  design      = inputs$design,
  threshold   = 3.0,
  formulas    = list("Condition x Time" = value ~ condition * time),
  output_file = NULL,
  min_cluster_size = 5
)
result
```

That call thresholds the stat map, finds clusters, labels each peak with its
atlas region and MNI coordinates, extracts mean signal per cluster, generates
time-course plots grouped by condition, formats a publication-ready coordinate
table, and renders orthographic brain slices at each peak.

Set `output_file = "report.pdf"` (or `.html`) to get a rendered document
instead of returning data.

## What you get back

`cluster_report()` returns a `cluster_report_result` object. It behaves like a
list -- you can extract components with `$` -- but it also has `print()`,
`summary()`, `plot()`, and `export_csv()` methods.

```{r summary}
summary(result)
```

## The MNI coordinate table

`result$mni_table` is a formatted table ready for publication:

```{r mni-table, results='asis'}
result$mni_table
```

## Cluster plots

The `plot()` method shows the time-course plot for a given cluster:

```{r plot-first, fig.cap = "Time course for the first cluster, split by condition."}
plot(result, which = 1)
```

You can also access plots directly:

```{r plot-direct, fig.cap = "Same plot, accessed directly."}
result$plots[["Condition x Time"]][[1]]
```

## How formulas control the plot

The formula does not fit a model -- your data is already fitted (betas,
contrasts, or summary scores). Instead, the formula controls **grouping and
layout**:

| Formula | X-axis | Color | Facet |
|---------|--------|-------|-------|
| `value ~ time` | time | -- | -- |
| `value ~ condition * time` | time | condition | -- |
| `value ~ group * condition * time` | time | condition | group |

The last RHS term is always the x-axis (or a column named `time` if present).
Additional terms map to color and facet in order.

Note that `+` and `*` are treated identically in this context -- both are
parsed with `all.vars()` to extract variable names. The formula operator does
not affect the plot.

You can include multiple formulas in one report:

```{r multi-formula, eval = FALSE}
cluster_report(
  ...,
  formulas = list(
    "Main Effect"  = value ~ time,
    "By Condition"  = value ~ condition * time
  )
)
```

Each formula gets its own subsection in the PDF, within each cluster's section.

## Exporting to CSV

For supplementary materials, export the cluster table, parcel overlaps, and
time-course data:

```{r export-csv, eval = FALSE}
export_csv(result, dir = "supplementary/")
```

This writes `cluster_report_clusters.csv`, `cluster_report_parcels.csv`, and
`cluster_report_timecourses.csv`.

## Data source types

`cluster_report()` accepts three types of backing data:

**4-D `NeuroVec`** (shown above) -- the most common case. One volume per row
of the design table.

**List of `NeuroVol` objects** -- useful when betas are separate files:

```{r list-source, eval = FALSE}
betas <- lapply(beta_files, neuroim2::read_vol)

cluster_report(
  stat_map = stat_map,
  data_source = betas,
  design = design,
  ...
)
```

**`nftab` (neurotabs)** -- the design is embedded in the nftab object, so no
separate `design` argument is needed:

```{r nftab-source, eval = FALSE}
cluster_report(
  stat_map = stat_map,
  data_source = my_nftab,
  ...
)
```

## Using the building blocks individually

`cluster_report()` is a convenience wrapper. You can call each step yourself
for more control:

```{r building-blocks}
# Step 1: Cluster extraction (the compute engine)
cluster_data <- compute_clusters(
  data_source = inputs$data_source,
  atlas = inputs$atlas,
  stat_map = inputs$stat_map,
  sample_table = inputs$design,
  threshold = 3.0,
  min_cluster_size = 5
)

# Step 2: Enrich with MNI coordinates and atlas labels
enriched <- enrich_cluster_table(
  cluster_table = cluster_data$cluster_table,
  stat_map = inputs$stat_map,
  atlas = inputs$atlas,
  max_clusters = 10
)
```

```{r building-blocks-extract}
# Step 3: Extract time-course data
tc_data <- extract_cluster_data(
  data_source = inputs$data_source,
  cluster_info = cluster_data,
  design = inputs$design
)
head(tc_data)
```

```{r building-blocks-plot, fig.cap = "Single-cluster plot built manually."}
# Step 4: Plot one cluster
plot_cluster_timecourse(
  data = tc_data[tc_data$cluster_id == enriched$cluster_id[1], ],
  formula = value ~ condition * time,
  cluster_label = "Manual plot: left FrontalA"
)
```

```{r building-blocks-table, results='asis'}
# Step 5: Format the MNI table
format_mni_table(enriched, style = "kable")
```

## Customisation options

**Brain slices:** Orthographic brain views are generated by default. Disable
with `brain_slices = FALSE` for faster runs, or change the color map:

```{r brain-options, eval = FALSE}
cluster_report(..., brain_slices = TRUE, brain_cmap = "viridis")
```

**Table style:** Choose between `"gt"` (default, best for HTML), `"flextable"`
(best for Word/PDF), or `"kable"` (lightweight, no extra dependencies):

```{r table-styles, eval = FALSE}
cluster_report(..., table_style = "flextable")
```

**Tail mode:** Focus on positive activations, negative, or both:

```{r tail-mode, eval = FALSE}
cluster_report(..., tail = "positive")   # only positive clusters
cluster_report(..., tail = "negative")   # only negative clusters
cluster_report(..., tail = "two_sided")  # both (default)
```

**Cluster cap:** Limit how many clusters appear in the report to keep it
manageable:

```{r max-clusters, eval = FALSE}
cluster_report(..., max_clusters = 10)
```

Clusters are sorted by peak statistic magnitude, so the strongest effects
appear first.

**Custom Rmd template:** Supply your own template for full control over the
report layout:

```{r custom-template, eval = FALSE}
cluster_report(..., template = "my_template.Rmd")
```

The template receives `params$report_data` -- see the bundled template at
`system.file("templates", "cluster_report.Rmd", package = "neuromosaic")`
for the expected structure.

## A note on statistical assumptions

This package assumes your statistical map has already been corrected for
multiple comparisons (e.g., via cluster-level FWE, voxel-level FDR, or
permutation testing). The `threshold` parameter controls cluster formation,
not statistical inference. If you pass an uncorrected map, the clusters and
their reported coordinates will be valid, but you should not interpret them
as statistically significant without appropriate correction.

## Next steps

- `?cluster_report` -- full parameter documentation
- `?compute_clusters` -- the compute engine underneath
- `?plot_cluster_timecourse` -- fine-grained control over individual plots
- `?format_mni_table` -- table formatting options
- `cluster_explorer()` -- launch the interactive Shiny explorer
