---
title: Exploring Ad Hoc Neurotabs Datasets
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Exploring Ad Hoc Neurotabs Datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)

has_neurotabs <- requireNamespace("neurotabs", quietly = TRUE)
has_explorer_deps <- has_neurotabs &&
  requireNamespace("shiny", quietly = TRUE) &&
  requireNamespace("DT", quietly = TRUE) &&
  requireNamespace("ggiraph", quietly = TRUE)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

Suppose you have one subject-level map per folder:

- `sub-01/maps/AUC.nii.gz`
- `sub-02/maps/AUC.nii.gz`
- `...`
- `sub-30/maps/AUC.nii.gz`

and a TSV file with `subject`, a continuous covariate `measure`, and a
two-level grouping variable `group`. You do not want to write a neurotabs
manifest by hand. You want to build an `nftab` directly from that table and
open the interactive explorer so that, after picking a cluster or parcel, the
signal plot shows `AUC ~ measure` with separate fitted lines for each `group`.

That workflow is:

1. Build an ad hoc `nftab` with `neurotabs::nf_from_table()`.
2. Pass that dataset to `nf_cluster_explorer()`.
3. In the app, set `Plot Variable = measure`, choose the `Grouped Overlay`
   plot plugin, and set its grouping variable to `group`.

`neuromosaic::nf_cluster_explorer()` uses `neurotabs::nf_design(ds)`
automatically, so the columns in your TSV become explorer controls without a
separate `design` argument.

## Create a nested AUC dataset

This example creates a temporary study directory with the same nested structure
you would have on disk. The group-level `stat_map` and atlas come from
`example_cluster_inputs()`, while the subject-level AUC volumes are written as
separate NIfTI files.

```{r make-observation-table}
inputs <- neuromosaic::example_cluster_inputs()
n_subjects <- 30L

obs <- tibble::tibble(
  subject = sprintf("sub-%02d", seq_len(n_subjects)),
  group = rep(c("control", "patient"), each = n_subjects / 2),
  measure = seq(-1.5, 1.5, length.out = n_subjects),
  auc_path = file.path(subject, "maps", "AUC.nii.gz")
)

head(obs, 4)
```

```{r write-auc-volumes, eval = has_neurotabs, include = FALSE}
study_dir <- file.path(tempdir(), "neuromosaic-auc-study")
dir.create(study_dir, recursive = TRUE, showWarnings = FALSE)

space_3d <- neuroim2::space(inputs$stat_map)
dims <- as.integer(dim(inputs$stat_map))[1:3]

for (i in seq_len(n_subjects)) {
  subject_dir <- file.path(study_dir, obs$subject[i], "maps")
  dir.create(subject_dir, recursive = TRUE, showWarnings = FALSE)

  arr <- array(stats::rnorm(prod(dims), sd = 0.05), dim = dims)
  arr[2:4, 2:4, 2:4] <- 0.4 + 0.9 * obs$measure[i] +
    ifelse(obs$group[i] == "patient", 0.6, -0.6) +
    stats::rnorm(27, sd = 0.08)
  arr[7:9, 7:9, 7:9] <- 0.2 - 0.5 * obs$measure[i] +
    stats::rnorm(27, sd = 0.08)

  neuroim2::write_vol(
    neuroim2::NeuroVol(arr, space = space_3d),
    file.path(study_dir, obs$auc_path[i])
  )
}

obs_path <- file.path(study_dir, "participants.tsv")
utils::write.table(obs, obs_path, sep = "\t", quote = FALSE, row.names = FALSE)

stopifnot(file.exists(obs_path))
stopifnot(all(file.exists(file.path(study_dir, obs$auc_path))))
```

## Build the `nftab` without a manifest

`nf_from_table()` is the ad hoc entry point in neurotabs. The `auc_path` column
stores relative file paths, and `root` tells neurotabs where those files live.
Here `axes = "subject"` keeps `group` and `measure` as row metadata rather than
observation axes.

```{r build-nftab, eval = has_neurotabs}
ds <- neurotabs::nf_from_table(
  observations = obs_path,
  feature = "AUC",
  locator_col = "auc_path",
  axes = "subject",
  space = "MNI152NLin2009cAsym",
  dataset_id = "auc-demo",
  root = study_dir
)

ds
```

```{r inspect-design, eval = has_neurotabs}
design_tbl <- neurotabs::nf_design(ds)
design_tbl[1:4, c("subject", "group", "measure")]
```

Those design columns are the bridge into `neuromosaic`:
`neuromosaic::nf_cluster_explorer()` reads `nf_design(ds)` and merges it into
the extracted cluster signal table.

## Launch the interactive explorer

`cluster_explorer()` expects both a volumetric atlas and a `surfatlas`. If you
already have a real surface atlas, use it. For a volume-only workflow, a small
label-only `surfatlas` is enough to enable the fallback parcel view used here.

```{r make-surfatlas}
surfatlas <- structure(
  list(
    ids = inputs$atlas$ids,
    labels = inputs$atlas$labels,
    hemi = inputs$atlas$hemi
  ),
  class = c("toy_surf", "surfatlas", "atlas")
)
```

```{r build-app, eval = has_explorer_deps}
app <- neuromosaic::nf_cluster_explorer(
  ds = ds,
  stat_map = inputs$stat_map,
  data_feature = "AUC",
  atlas = inputs$atlas,
  surfatlas = surfatlas,
  threshold = 3,
  min_cluster_size = 5
)

inherits(app, "shiny.appobj")
```

In an interactive R session, run `app` or `shiny::runApp(app)`.

Once the app is open:

1. Click a parcel in the brain view or select a cluster from the table.
2. Set `Plot Variable` to `measure`.
3. Open `Signal Controls`.
4. Leave `Analysis Plugin` as `None (raw signal)` unless you want to transform
   the extracted ROI table first.
5. Set `Plot Plugin` to `Grouped Overlay`.
6. Set `Grouping variable` to `group`.
7. Leave `Collapse By` empty unless you want to average repeated rows first.

The signal panel will show subject-level AUC values for the selected cluster,
with points colored by `group` and separate linear fits overlaid for each
group. If you select multiple clusters, the explorer facets them so each panel
keeps the same `measure`/`group` mapping.

## Plot plugins are the extension point

`Grouped Overlay` is only one bundled plot plugin. The explorer always extracts
the same ROI-by-sample signal table; plot plugins decide how that table is
rendered.

A plot plugin receives:

- `data`: the selected signal table, including `signal`, `cluster_id`, and the design columns carried over from the neurotab.
- `params`: values collected from the plugin's own UI controls.
- `context`: the current plot settings, including `x_var`, `collapse_vars`, and the available continuous and categorical columns.
- `interactive`: whether the explorer is building a static `ggplot` or an interactive `girafe` widget.

A plugin can return a `ggplot`, a `girafe`, or a list with `plot` and optional
`diagnostics`. That means the `measure + group` example is not hard-coded into
the explorer. It is just one way to render the extracted table.

```{r custom-plot-plugin, eval = has_explorer_deps}
scatter_lm_plugin <- list(
  id = "scatter_lm",
  label = "Scatter + LM",
  param_defs = list(
    list(
      name = "colour_var",
      label = "Colour variable",
      type = "select",
      default = "",
      choices = function(context) {
        cols <- context$categorical_columns
        if (is.null(cols)) cols <- character(0)
        c("None" = "", stats::setNames(cols, cols))
      }
    )
  ),
  render = function(data, params, context, interactive) {
    plot_data <- data
    plot_data$.x <- plot_data[[context$x_var]]
    colour_var <- params$colour_var
    if (is.null(colour_var)) colour_var <- ""

    if (nzchar(colour_var)) {
      plot_data$.colour <- plot_data[[colour_var]]
      p <- ggplot2::ggplot(
        plot_data,
        ggplot2::aes(x = .x, y = signal, colour = .colour)
      ) +
        ggplot2::geom_point() +
        ggplot2::geom_smooth(method = "lm", se = FALSE)
    } else {
      p <- ggplot2::ggplot(plot_data, ggplot2::aes(x = .x, y = signal)) +
        ggplot2::geom_point() +
        ggplot2::geom_smooth(method = "lm", se = FALSE)
    }

    if (dplyr::n_distinct(plot_data$cluster_id) > 1) {
      p <- p + ggplot2::facet_wrap(~ cluster_id, scales = "free_y")
    }

    p + ggplot2::theme_minimal()
  }
)

app_custom <- neuromosaic::nf_cluster_explorer(
  ds = ds,
  stat_map = inputs$stat_map,
  data_feature = "AUC",
  atlas = inputs$atlas,
  surfatlas = surfatlas,
  threshold = 3,
  min_cluster_size = 5,
  plot_plugins = list(scatter_lm = scatter_lm_plugin),
  default_plot_plugin = "scatter_lm"
)

inherits(app_custom, "shiny.appobj")
```

When that app is open:

1. Pick a cluster or parcel.
2. Set `Plot Variable` to `measure`.
3. Open `Signal Controls`.
4. Leave `Plot Plugin` as `Scatter + LM`.
5. Set `Colour variable` to `group`.

The result is still driven by the same extracted ROI table, but now the signal
panel is using your custom renderer instead of one of the built-in plugins.

## Direct template for your own study

If your data already lives on disk, the only part you need to change is the
table and paths:

```{r your-study-template, eval = FALSE}
obs <- utils::read.delim("~/study/participants.tsv")

ds <- neurotabs::nf_from_table(
  observations = obs,
  feature = "AUC",
  locator_col = "auc_path",
  axes = "subject",
  space = "MNI152NLin2009cAsym",
  root = "~/study"
)

app <- neuromosaic::nf_cluster_explorer(
  ds = ds,
  stat_map = stat_map,
  data_feature = "AUC",
  atlas = atlas,
  surfatlas = surfatlas
)
```

For larger datasets, `nf_cluster_explorer(..., strategy = "lazy")` avoids
stacking every subject volume up front and samples from the NIfTI files on
demand.
